Co-methylation review

### 2015-04-21 Rmarkdown/comethylation_review.Rmd
Complete my edits to Co-methylation

__TODO: Review co-methylation in "Bell, J. T. et al. DNA methylation patterns associate with genetic and gene expression variation in HapMap cell lines. Genome Biol 12, R10 (2011).__"

@@

@@

Between-sample co-methylation is somewhat analogous to the idea of linkage disequilibrium between genetic loci since it is a locus-specific correlation measured at the population level.

@@

An example of between-sample co-methylation is given by \citet{Liu:dy}. The authors compute the correlation of pairs of $\beta$-values across samples to identify clusters of correlated CpGs and relate methylation changes at these clusters to nearby SNP genotypes. The resolution of their data is limited since it uses Illumina HumanMethylation450 array, but they conclude that "DNA methylation is correlated over regions with a median length of 274 bp from our data set". They also note that, "methylation codependency [co-methylation] itself varies across the genome", as is to be expected. No software is provided to implement the method.

A simple way to estimate between-sample co-methylation is to create a matrix of $\beta$-values, $\bm{\beta}$, where the $(i, j)$ entry is $\beta_{i, j}$. Then, compute $R = cor(\bm{\beta})$, whose $(k, l)$ entry is $cor(\bm{\beta_{., k}}, \bm{\beta_{., l}})$, where $\bm{\beta_{., k}}$ is the column vector of $\beta$-values for the $k^{th}$ methylation loci. This is what the R package, `coMET`   ([http://bioconductor.org/packages/devel/bioc/html/coMET.html](http://bioconductor.org/packages/devel/bioc/html/coMET.html)), does to estimate the between-sample co-methylation of Illumina HumanMethylation450 array data.

@@

It has been difficult to properly account for co-methylation in downstream analyses due to the variety of definitions, the varying rigour and reproducibility of the published analyses and the lack of software to estimate co-methylation.. Ideally, co-methylation would not only be _accounted for_ in downstream analyses but _exploited_ to leverage the power of the correlation structure of DNA methylation data. Some authors have already sought to do this.

For example, the `bsseq` R/Bioconductor package for identifying differentially methylated regions from WGBS data \citep{Hansen:2011gu, Hansen:2012gr} cite \citet{Eckhardt:2006gh} as evidence that methylation levels at nearby CpGs are correlated. `bsseq` uses local binomial likelihood smoothing to exploit the spatial correlation of $\beta$-values. Specifically, by assuming that CpGs close to one another have similar methylation levels, the smoothed estimates of the $\beta$-values are, on average, more accurate than the raw $\beta$-values. This is particularly powerful for experiments with low sequencing coverage since the error in estimating the raw $\beta$-values is inversely proportional to the sequencing coverage and smoothing reduces this error.

The R/Bioconductor package `BiSeq` \citep{Hebestreit:2013ko} also smooths the $\beta$-values, albeit slightly differently to `bsseq` since it is designed for identifying differentially methylated regions from RRBS data. The authors also cite \citet{Eckhardt:2006gh} as justification for smoothing the $\beta$-values.

To summarise, both `bsseq` and `BiSeq` smooth $\beta$-values on a per-sample basis and are motivated by an analysis of within-sample co-methylation. Whenever using statistical smoothing, an important parameter is the choice of window size over which to smooth, also known as the _bandwidth_ of the smoother. The default window size in `bsseq` is defined as one that contains at least 70 CpGs and is at least 2000 bp wide. `BiSeq` uses a much smaller default window size of 80 bp, a decision that is again driven by its focus on RRBS data which has a higher CpG density than WGBS data. Ideally, a data-driven window size would be used, one driven by the strength of co-methylation in the region, but this is difficult to implement in practice. A simpler but potentially useful hybrid would be to define different window sizes for genomic features with generally different co-methylation structures, e.g., CpG islands vs. the rest of the genome (see Chapter \ref{chap:co-methylation}).

`BiSeq` also estimates and tries to account for the spatial correlation between test statistics of differential methylation at neighbouring loci. This is not the same as looking for correlations between the $\beta$-values, but is rather based on looking for correlations amongst the _differences_ in $\beta$-values, since these differences are what the test statistics of differential methylation are based upon. Specifically, `BiSeq` fits a semivariogram to the test statistics of differentially methylated CpGs. The correlations estimated using the semivariogram are in turn used in estimates of the standard deviations when identifying differentially methylated regions.

More generally, methods such as `SLIM` \citep{Wang:2011cw} and `comb-p` \citep{Pedersen:2012vl} attempt to account for the correlation amongst test statistics in genomics data. They do this by analysing the distribution of P-values rather than the distribution of the test statistics themselves or the underlying data. This makes these methods quite general but more application-specific methods may be more powerful. `comb-p` has been used in the analysis of DNA methylation data from the Illumina microarray platforms \citep{Pedersen:2012vl}.

@@

\citet{Capra:2014hx} also make use of between-sample co-methylation, albeit a specialised form of it. They study the dynamics of DNA methylation in differentiating cell lineages. For each locus there is a correlation between its methylation state in the precursor cell and its dependent cell types, which they call "vertical" correlations in contrast to the "horizontal" correlations of within-sample co-methylation. \citet{Capra:2014hx} are bringing well-established methods for the study of DNA sequence evolution to the study of DNA methylation dynamics. In this specialised setting, where "between-samples" means "between samples in a lineage", they show that a method based on "vertical" correlations is better at imputing the methylation level of CpGs than a method based on "horizontal" within-sample co-methylation.

@@

While it has long been observed that DNA methylation at nearby loci is highly correlated, pinning down exactly what this means, and its implications, is surprisingly difficult. As I have shown, there are problems with most, if not all, previous analyses. These problems range from a lack of methodological detail and limited data, through to serious flaws in the methodology or its implementation. Moreover, despite numerous papers including an analysis of co-methylation, none of these have includes software to reproduce their method, further complicating the assessment of these methods.

>>>>

@@

\subsection{\citet{Ball:2009kh}}

\citet{Ball:2009kh} use bisulfite padlock probes with $36$ bp single-end reads to profile CpG methylation at approximately $7,000$ CpGs in an Epstein-Barr virus transformed B-lymphocyte cell line. As part of their study, \citeauthor{Ball:2009kh} sought to investigate whether the co-methylation reported by \citet{Eckhardt:2006gh} (Section \label{sec:eckhardt}) occurs at the single-molecule level.

To do so, \citeauthor{Ball:2009kh} took all reads containing multiple CpGs and mapping to positions with intermediate $\beta$-values ($0.2 < \beta < 0.8$) with at least $100\times$ sequencing coverage. From these reads they computed the correlation of within-read methylation states. \citeauthor{Ball:2009kh} found that the distribution of these correlations is centred around $0.5$ with very few values below zero, evidence that methylation levels at CpGs on the same DNA fragment "are generally positively correlated".

The obvious limitation to this analysis is the small number of CpG pairs interrogated by the assay. The short reads also only allow for the exploration of within-fragment co-methylation at very close CpGs and \citeauthor{Ball:2009kh} do not explore the effect of distance between CpGs ($IPD$) on co-methylation.

@@

@@

\citet{Bell:2011fm} identify evidence of between-sample co-methylation using the Illumina 27k microarray to measure DNA methylation in lymphoblastoid cell lines from 77 HapMap Yoruba individuals. Of note, they found that between-sample co-methylation decays as a function of the distance between CpGs ($IPD$) and is stronger within CpG islands than outside of CpG islands. However, it must be recognised that the Illumina 27k microarray measures DNA methylation at less than $1\%$ of all CpGs as is biased towards CpG islands.

Another example of between-sample co-methylation is given by \citet{Liu:dy}. The authors compute the correlation of pairs of $\beta$-values between samples in three studies ($n = 247, n = 91$ and $n= 305$, respectively) to identify clusters of correlated CpGs and relate methylation changes at these clusters to nearby SNP genotypes. The resolution of their data is better than \citet{Bell:2011fm} since they use the Illumina 450k microarray, however it still only captures less than $2\%$ of all CpGs in the genome. \citeauthor{Liu:dy} conclude that "DNA methylation is correlated over regions with a median length of 274 bp from our data set". They also note that, "methylation codependency [co-methylation] itself varies across the genome", as is to be expected.

A simple way to estimate between-sample co-methylation is to create a matrix of $\beta$-values, $\bm{\beta}$, where the $(i, j)$ entry is $\beta_{i, j}$. Then, compute $R = cor(\bm{\beta})$, whose $(k, l)$ entry is $cor(\bm{\beta_{., k}}, \bm{\beta_{., l}})$, where $\bm{\beta_{., k}}$ is the column vector of $\beta$-values for the $k^{th}$ methylation loci. This is what the R package, `coMET`   ([http://bioconductor.org/packages/devel/bioc/html/coMET.html](http://bioconductor.org/packages/devel/bioc/html/coMET.html)), does to estimate the between-sample co-methylation of Illumina HumanMethylation450 array data. As far as I am aware, `coMET` is the only software specifically designed to estimate between-sample co-methylation and integrate this with other genomic annotations.

@@

The variety of definitions, rigour, and reproducibility of the published analyses of co-methylation, not to mention the lack of software implementing proposed methods, has made it difficult to properly account for co-methylatoin in downstream analyses. Ideally, co-methylation would not only be _accounted for_ in downstream analyses but _exploited_ to leverage the power of the correlation structure of DNA methylation data. In fact, some authors have already sought to do this.

For example, the R/Bioconductor packages `bsseq` \citep{Hansen:2011gu, Hansen:2012gr} and `BiSeq` \citep{Hebestreit:2013ko} both cite \citet{Eckhardt:2006gh} as providing evidence of co-methylation, which can be exploited by spatial smoothing of the $\beta$-values prior to testing for differential methylation. Specifically, by assuming that CpGs close to one another have similar methylation levels, the smoothed $\beta$-values are, on average, more accurate than the raw $\beta$-values. Both `bsseq` and `BiSeq` smooth $\beta$-values on a per-sample basis.

Smoothing is particularly powerful for experiments with low sequencing coverage since the error in estimating the raw $\beta$-values is inversely proportional to the sequencing coverage and smoothing reduces this error. An important parameter when smoothing is the choice of window size over which to smooth, also known as the _bandwidth_ of the smoother. The default window size in `bsseq` is defined as one that contains at least 70 CpGs and is at least 2000 bp wide. `BiSeq` uses a much smaller default window size of 80 bp, a decision that is  driven by its focus on RRBS data which has a higher CpG density than whole-genome bisulfite-sequencing data. Ideally, a data-driven bandwidth would be used, one driven by the strength of co-methylation in the region, but this is difficult to implement in practice. A simpler but potentially useful hybrid would be to define different window sizes for genomic features with generally different co-methylation structures, e.g., CpG islands vs. the rest of the genome (see Chapter \ref{chap:co-methylation}) (__TODO: Revisit in light of updated results in Co-methylation chapter__).

In addition to smoothing the $\beta$-values, `BiSeq` also estimates and tries to account for the spatial correlation of test statistics when testing for differential methylation at neighbouring loci. This is not the same as looking for correlations between the $\beta$-values, but is rather based on looking for correlations amongst the _differences_ in $\beta$-values, since these differences are what the test statistics of differential methylation are based upon. Specifically, `BiSeq` fits a semivariogram to the test statistics of differentially methylated CpGs. The correlations estimated using the semivariogram are in turn used in estimates of the standard deviations when identifying differentially methylated regions.

More generally, methods such as `SLIM` \citep{Wang:2011cw} and `comb-p` \citep{Pedersen:2012vl} attempt to account for the correlation amongst test statistics in genomics data. They do this by analysing the distribution of P-values rather than the distribution of the test statistics themselves or the underlying data. This makes these methods quite general but application-specific methods may be more powerful. `comb-p` has been used in the analysis of DNA methylation data from the Illumina microarray platforms \citep{Pedersen:2012vl}.

@@

\citet{Capra:2014hx} also make use of between-sample co-methylation, albeit a specialised form of it, to study the dynamics of DNA methylation in differentiating cell lineages. For each locus there is a correlation between its methylation state in the precursor cell and its dependent cell types, which they call "vertical" correlations (in contrast to the "horizontal" correlations of within-sample co-methylation). In this specialised setting, where 'between-samples' means "between samples in a lineage", they show that a method based on "vertical" correlations is better at imputing the methylation level of CpGs than a method based on "horizontal" within-sample co-methylation.

@@

While it has long been observed that DNA methylation at nearby loci is highly correlated, pinning down exactly what this means, and its implications, is surprisingly difficult. As I have shown, there are problems and limitations with most, if not all, previous analyses. These problems range from a lack of methodological detail and limited data, through to serious flaws in the methodology or its implementation. Moreover, despite numerous papers including an analysis of co-methylation, none of these have includes software to reproduce their method, further complicating the assessment of these methods.
